Leadership across cultures and contexts: an exploratory analysis

### 2020-05-18 leadership_across_cultures_contexts.Rmd
refer back to Trobriand use of sorcerers.
Read throuhg with minor revisions.
Added some discussion in SI on bias.

Updated typo in variable chart.

Despite decades of research and theorizing on leadership in numerous social science disciplines, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

Anthropology, on the other hand, has been criticized for exoticizing and essentializing diverse "Others" as part of its colonial history [@said1979; @abu2008; @reyes-garcia_small-scale_2017]. @abu2008 [p. 470], for instance, notes that

@@

There is potentially as much diversity within non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking the nature of leadership and followership across social contexts and levels of social organization, such as among kin, economic, and military groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal, and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with increased sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2019]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

@@

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. Sex, however, is confounded with factors, such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social political influence and maintain control through a variety of mechanisms including military force, ideological systems, and control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts between individuals can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and social ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

@@

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such as a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) gain other social or reproductive benefits by signaling their high quality as a mate and social partner [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019], even if 'just for a while.'

@@

The text records were extracted from the 60-culture Probability Sample Files (PSF) of the eHRAF [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling of one culture with high quality ethnographic coverage from each of 60 culture regions. The database of text records was designed to capture a wide sample of the ethnographic record of leadership across social contexts and was built using an advanced search including 16 OCM codes related to leadership and seven key words describing leaders or followers. See @garfield_evolutionary_2019 for additional details on the ethnographic sample and methodology.

@@

In contrast to the qualities and functions dimensions, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on the theoretical literature. For both followers and leaders, we coded evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). Some of the costs and benefits (e.g., mating) were similar to leader qualities (e.g., polygyny). The distinction is that costs and benefits were transactional whereas qualities were descriptive (e.g., leaders obtain wives vs. having wives).

@@

Each document from each culture was coded for date of publication and the sex of the ethnographer(s). For additional details on ethnographic sampling and creation of the database of leadership text records, see @garfield_evolutionary_2019.

@@

Our third goal was to assess the extent to which each of the dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. We also wanted to investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions. For these analyses, we often used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful with the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. See the SI for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Subsistence strategy and group structure were also confounded. Hunter-gatherer text records often referred to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely referred to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other hand, rarely referring to residential leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, political groups at the community level were the most frequent type, with the exception of agriculturalists where supracommunity political groups were the most frequent type. Leadership beyond the community level was relatively rare outside of agricultural populations and was especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

The majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with the remaining records not sex-specific. See Table S\@ref(tab:fourfactorsumm). There were female leaders in almost all types of groups, however, except military and religious. Female leaders did tend to lead at the kin group and other residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, for instance, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%). See Table S\@ref(tab:sexsubtable).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by date of publication of the source ethnographic texts, female gender of any author or co-author, or total pages of ethnography from each culture. Dates of publication ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Figure S\@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`, and this was weakly correlated with the number of text records on leadership retrieved for a particular culture (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). See Figure S\@ref(fig:eHRAFpages).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary leadership dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but recent publications were significantly more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information (SI) for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support, as described in the Methods section, and then computed the proportion of cultures with at least one text record providing support for that dimension. See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

@@

```{r benefitcostratio, fig.width=10, fig.cap="The odds ratios of finding evidence of a benefit vs. a cost in a text record for leaders and followers, by each dimension of benefits and costs. Odds ratios estimated using a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. Separate models were fit for leaders and followers. X-axis is on a log scale."}
ggsave(file='Figures/plot_bc_OR.pdf', plot_bc_OR, width = 9, height = 5)
plot_bc_OR
```

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:highstatusplot)A.

To determine which leader qualities and functions were associated with evidence for high status leaders, we fit an elasticnet logistic model of *High status* as a function of all quality and function dimensions, (with the lasso penalty, $\alpha = 1$). The non-zero coefficients indicated that *Punishment*, the only identified function, was a negative predictor, whereas
<!-- all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and  -->
*Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several other qualities, were positive predictors. See Figure \@ref(fig:highstatusplot)B.

@@

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of competencies, a single feature that included at least two important theoretical elements: prestige and dominance. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between features, after aggregating leadership dimensions within features (see Figure \@ref(fig:featurecor)).

>>>>

Despite decades of research and theorizing on leadership in numerous disciplines, to our knowledge there is no systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

Anthropology, on the other hand, has been criticized for exoticizing and essentializing diverse "Others" as part of its colonial history [@said1979; @abu2008]. @abu2008 [p. 470], for instance, notes that

@@

There is potentially more diversity within non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking leadership dynamics across contexts and levels of social organization, such as among kin and social groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with increased sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2019]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

@@

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideological systems, and control of information [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts between individuals can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

@@

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such as a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers; or (4) gain other social or reproductive benefits by signaling their high qualities [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019], even if 'just for a while.'

@@

The text records were extracted from the 60-culture Probability Sample Files (PSF) of the eHRAF [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling of one culture with high quality ethnographic coverage from each of 60 culture regions. The database of text records was designed to capture a wide sample of the ethnographic record of leadership across social contexts and was built using an advanced search including 16 OCM codes related to leadership and seven key words describing leaders or followers. See @garfield_evolutionary_2019 for additional details on the ethnographic sample.

@@

In contrast to the qualities and functions dimensions, operationalizations of the costs and benefits of leadership were developed *a priori* based on theoretical literature. For both followers and leaders, we coded evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). Some of the costs and benefits (e.g., mating) were similar to leader qualities (e.g., polygyny). The distinction is that costs and benefits were transactional whereas qualities were descriptive (e.g., leaders obtain wives vs. having wives).

@@

Each document was coded for publication date and the ethnographer(s) sex (some metadata collected/confirmed with @mehr_natural_2016). For additional details on ethnographic sampling and creation of the text record database, see @garfield_evolutionary_2019.

@@

Our third goal was to assess the extent to which each of the dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. We also wanted to investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions. For these analyses, we often used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful when the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. See the supplementary information (SI) for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Subsistence strategy and group structure were also confounded. Hunter-gatherer text records often referred to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely referred to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other hand, rarely referred to residential leadership (`r agri_residential_pct`%) but often referred to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, political groups at the community level were the most frequent context, with the exception of agriculturalists where supracommunity political groups were the most frequent. Leadership beyond the community level was relatively rare outside of agricultural populations and was especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

The majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with the remaining records not sex-specific. See Table S\@ref(tab:fourfactorsumm). There were female leaders in almost all contexts, however, except military and religious. Female leaders did tend to lead within kin and residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, for instance, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%). See Table S\@ref(tab:sexsubtable).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by publication date of source ethnographic texts, female gender of any author or co-author, or total pages of ethnography from each culture. Publication dates ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Figure S\@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`, and this was weakly correlated with the number of text records on leadership retrieved for a particular culture (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). See Figure S\@ref(fig:eHRAFpages).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary leadership dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but recent publications were significantly more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and SI for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (as described in the Methods section). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing cooperative activities, and providing counsel or direction to followers. In about half of cultures sampled here leaders punished group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

@@

```{r benefitcostratio, fig.width=10, fig.cap="Odds ratios of finding evidence of a benefit vs. a cost in a text record for leaders and followers, by each dimension of benefits and costs. Odds ratios estimated using a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. Separate models were fit for leaders and followers. X-axis is on a log scale."}
ggsave(file='Figures/plot_bc_OR.pdf', plot_bc_OR, width = 9, height = 5)
plot_bc_OR
```

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:highstatusplot)A.

To determine which leader qualitiy and function dimensions were associated with evidence for high status leaders, we fit an elasticnet logistic model of *High status* as a function of all quality and function dimensions, (with the lasso penalty, $\alpha = 1$). The non-zero coefficients indicated that *Punishment*, the only identified function, was a negative predictor, whereas
<!-- all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and  -->
*Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several other qualities, were positive predictors. See Figure \@ref(fig:highstatusplot)B.

@@

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of *Competencies*, a single feature that included at least two important theoretical elements: dominance and prestige. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between features, after aggregating leadership dimensions (see Figure \@ref(fig:featurecor)).

We then fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting the `r nrow(feature_models_aic)` models whose AIC values were at least two points less than the intercept-only models (only *Cultural conformity* was dropped). As with our leader dimension models, we then dropped each factor variable in turn and compared the change in AIC. The large increases for group context and leader sex indicated that including these factors substantially improved fit. See Figure \@ref(fig:heatmapdrop1)B. We then compared the levels of each factor that improved fit for each feature variable. Female leaders were associated with *Social, reproductive, and material success*, and male leaders with *(Prosociality*. Military groups were associated the *Organize* feature, religious groups with *Prosociality* and *Competencies*, and kin groups with *Strategize* and *Prosocial competencies*. See Figures S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).
